---
title: "Multivariate Adaptive Regression Splines"
author: "Zenith (Robert) Tandava Yurkovich, Valeriya Kistrina"
date: "2023-04-16"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### MARS package GitHub Repo link:

#### <https://github.com/CPSCValeriya/STAT_360_mars_Project/tree/main/mars>

#### Please use: install_github(repo="CPSCValeriya/STAT_360_mars_Project/mars")

**We recommend the viewer to run rm(list=ls()) to clear their environment before running our package. Steps we have followed have been clearing the environment and running load_all()**

## Description

mars, which stands for Multivariate Adaptive Regression Spline, is used to apply a regression model to a given set of data. In doing so, the function analyzes the dataset and creates numerous knots: points at which the fitted model will change.

## Usage

```{r eval=FALSE}
#With default parameters
mars(formula, data, control=mars.control())

#With specified arguments
mars(formula, data, control=mars.control(Mmax=3, d=2, trace=TRUE))
```

## Arguments

-   **formula**: an object of class "formula" (or one that can be coerced to that class): a symbolic description of the model to be fitted

-   **data:** a data frame that contains the data on which we wish to apply the mars() function

-   **control**: an object of class mars.control. If unspecified, will be defaulted to a mars.control object with Mmax = 2, and d = 3 (see Details for explanation of these parameters)

## Details

Although many other R functions (such as lm() and glm()) can be used to fit a regression models onto datasets, mars() differs in several ways. Most notably, mars() is non-parametric unlike the aforementioned functions which return models with predictor and response variables. Instead, mars() analyzes the dataset and creates numerous break points at which, the model changes hence, adaptive splines. As such, the model produced by the mars() function is piecewise, unlike the outputs of the lm() and glm() functions. Nevertheless, there do exist important similarities between other model fitting functions and mars(). Notably, the formula input is present in mars, lm, and glm, which means that all of the attributes of the formula class can be applied to the formula argument for mars (see ?lm or ?glm)

The points at which the function splits and create a new spline are determined by partitioning the dataset to minimize a lack-of-fit (LOF) function (see Friedman paper for more details).

The knots produced by the mars function can be influenced by the user via the control argument. The control argument, which is of class mars.control, has two main components: Mmax, and d.

-   **Mmax:** a positive even integer argument that specifies the maximum number of splines that our output model can have

-   **d**: a parameter that affects the LOF function and thereby affects how many knots are produced. In general, a larger d value will result in a smaller number of knots in the output of the mars function

## **Value**

The mars() function returns an object of class mars. The mars object is a list containing at least the following components (as well as the components of its parent class lm which are not all listed):

-   **y:** the formula's response variable's values

-   **B:** a matrix of values by which the coefficients will be multiplied (see also coefficients)

-   **Bfuncs:** a list containing the knot points

-   **call:** the call of the mars function with the given inputs

-   **formula:** the formula argument of the mars function

-   **x_names:** a list of all of the predictor variable names

-   **coefficients**: a vector of coefficient values, which will be multiplied by the B matrix

-   **residuals:** the response minus the fitted values

-   **rank**: the numerical rank of the fitted model

-   **fitted.values**: the fitted model values

-   **df.residual:** the residual degrees of freedom

-   **xlevels**: the levels of the response variables (if applicable)

-   **terms**: general information about the model, its coefficients, and its response

-   **model**: a matrix which is the column vector y column bound to the matrix B

## Authors

Valeriya (Valerie) Kistrina, Christine Chuong, Zenith (Robert) Tandava Yurkovich

## References

Friedman, J. H. (1991). Multivariate Adaptive Regression Splines, The Annals of Statistics, Vol. 19, No. 1, pp. 1-67

Milborrow, S. H., Hastie, T., Tibshirani, R., Miller, A., & Lumley, T. (2021). earth: Multivariate Adaptive Regression Splines, R package version 5.3.1.

Lin, W. (2023). Lecture Material.

## See Also

-   **print.mars:** prints out a user-friendly output of the mars object details (includes the coefficients, basis functions, product and call)

-   **summary.mars:** provides a detailed summary of the mars model object (includes the coefficients, basis functions, residuals, additional statistics and other)

-   **anova.mars:** prints the Analysis of Variance Table (ANOVA) and details on the corresponding mars object

-   **predict.mars:** generates a prediction, obtaining estimated values of the response variable given a different dataset and mars model object

-   **plot.mars:** generates four different plots based on the provided mars object (Residuals vs. Fitted, Quantile-Quantile, Scale-Location, and Residuals vs. Leverage)

## Examples

Below is a collection of examples on how to use the mars function and methods written for its output:

#### Example 1.1: Utilizing the mars() function with Boston city data from the MASS package

-   Please note: an increase in Mmax increases the time required to return the mars output

```{r}
library(MASS)

#Using the mars() function with default control values
Boston.mars1 = mars(crim~age+nox+rm+dis+tax, Boston, control = mars.control())

#Running mars() function but changing the control argument
#Mmax is increased (2 -> 10) and d is decreased (3 -> 2)
#These two will lead to more knots than Boston.mars1
#See use of trace = TRUE to print progress details 
Boston.mars2 = mars(crim~age+nox+rm+dis+tax, Boston,
                    control = mars.control(Mmax = 6, d = 2, trace = TRUE))
```

#### Example 1.2: Utilizing the mars() object methods (print, summary, anova)

-   See Example 2 for predict function and Example 3 for plot function

```{r}
#Method 1: print
cat("Boston.mars1 object -------------------------------\n\n");
print(Boston.mars1) # Shows 2 B coefficients
cat("\nBoston.mars1 object -------------------------------\n\n");
print(Boston.mars2) # Shows 5 B coefficients
```

```{r}
#Method 2: summary
cat("Boston.mars1 object -------------------------------\n\n");
summary(Boston.mars1)
cat("\nBoston.mars2 object -------------------------------\n\n");
summary(Boston.mars2)
```

```{r}
#Method 3: anova
cat("Boston.mars1 object -------------------------------\n\n");
anova(Boston.mars1)
cat("\nBoston.mars2 object -------------------------------\n\n");
anova(Boston.mars2)
```

#### Example 2: Utilizing the predict method with the Iris data set 

```{r}
#Calling mars with default parameters and applying predict.mars
Iris.Mars <-mars(Sepal.Length~Sepal.Width+Petal.Length+Petal.Width,
                 data=iris,control=mars.control())

#Returns the "fitted.values" component of Iris.mars since data set is unchanged
prediction_iris_no_noise = predict(Iris.Mars)
print(prediction_iris_no_noise)
```

```{r}
#Altering the data set to see the how the prediction changes by adding noise terms to the explanatory variables
iris2 = iris
set.seed(123)
iris2$Sepal.Width = round(iris$Sepal.Width + rnorm(150),1)
iris2$Petal.Length = round(iris$Petal.Length + rnorm(150),1)
iris2$Petal.Width = round(iris$Petal.Width + rnorm(150, sd = 0.1)+0.2,1)

#New predicted values on dataset with noise
prediction_iris_with_noise = predict(Iris.Mars,iris2)
print(prediction_iris_with_noise)
```

#### Example 3.1: Using default plot on mars objects with Cars93 data from the MASS package

-   Please note: Hit \<Return\> to see the plots

```{r}
#Calling mars with specified parameters (Mmax = 6, d = 3, trace = FALSE)\n")
Car.mars = mars(Rev.per.mile~EngineSize+Horsepower+RPM+as.factor(Origin)+
                  Fuel.tank.capacity+MPG.city+MPG.highway,
                data = Cars93, mars.control(Mmax = 6, d = 3, trace = FALSE))

#Plotting default plot (cycles through the 4 plots that plot.mars produces)
cat("\nplot with default arguments\n")
plot(Car.mars)
```

#### Example 3.2: Plot with additional arguments on mars objects with Cars93 data 

```{r}
#Additional parameters that can be altered include: which, col, numpoints

#which: plots can be done individually by specifying their index
#'1 = "CDF" plot of absolute value of residuals
#'2 = "Residual vs Fitted" plot
#'3 = "Normal Q-Q" plot
#'4 = "Scale-Location" plot

#Specify '1' for the Cumulative Distribution plot
cat("\n1= Cumulative Distribution plot\n")
plot(Car.mars, which = 1)

#Specify '2' for the Residuals vs. Fitted plot
cat("\n2= Residuals vs. Fitted plot\n")
plot(Car.mars, which = 2)

#Specify '3' for the Normal Q-Q plot
cat("\n3 = Normal Q-Q plot\n")
plot(Car.mars, which = 3)

#Specify '4' for the Scale-Location plot
cat("\n4 = Scale-Location plot\n")
plot(Car.mars, which = 4)
```

```{r}
#col: specify a colour of the points. Purple by default.
plot(Car.mars, col="blue")
```

```{r}
#numpoints: how many of the largest absolute value points to plot
plot(Car.mars, numpoints=50)
```

```{r}
#A combination of parameters can be specified:
plot(Car.mars, which=c(1,2,3),col="orange",numpoints=40)
```
